---
title: "Standard errors that respect the ESS sample design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standard errors that respect the ESS sample design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

> **User story.** As a quantitative researcher submitting for publication, I need
> confidence intervals and tests that account for the ESS's clustered, stratified
> samples, so that I do not report significance that rests on an assumption of
> simple random sampling that the ESS does not meet.

This vignette uses **survey**, which is a `Suggests` dependency — install it
separately. Catalogue output is real; results from survey data are shown as code
only.

```{r setup}
library(essurvey2)
library(data.table)
library(survey)
```

## Why the default standard errors are wrong

No ESS country draws a simple random sample of individuals. Samples are
stratified by region and often by settlement size, and in most countries drawn in
clusters — addresses within sampled areas. Respondents within a cluster resemble
each other, so each additional interview in a cluster adds less information than
an independent one would. `lm()`, `glm()` and `t.test()` assume otherwise, and
the usual result is standard errors that are too small, intervals that are too
narrow, and *p*-values that are too generous. Weights alone do not repair this:
they correct the point estimate, not its variance.

The fix needs three things from the data: the primary sampling unit, the stratum,
and a weight. Whether the ESS publishes them depends on the round, which is the
first thing to establish.

## Where the design variables live

From round 9, they are in the integrated file, in their own variable group:

```{r}
ess_variables(11)[group_name == "sampling", .(variable_name, variable_label)]
#>    variable_name        variable_label
#>           <char>                <char>
#> 1:        domain       Sampling domain
#> 2:          prob  Sampling probability
#> 3:       stratum      Sampling stratum
#> 4:           psu Primary sampling unit
```

Before round 9 they are not:

```{r}
c("psu", "stratum", "prob", "domain") %in% ess_variables(7)$variable_name
#> [1] FALSE FALSE FALSE FALSE
```

Checking every round at once:

```{r}
design_vars <- c("psu", "stratum", "prob")

rbindlist(lapply(1:11, function(r) {
  present <- ess_variables(r)$variable_name
  data.table(round = r, in_integrated = all(design_vars %in% present))
}))
#>     round in_integrated
#>     <int>        <lgcl>
#>  1:     1         FALSE
#> ...
#>  8:     8         FALSE
#>  9:     9          TRUE
#> 10:    10          TRUE
#> 11:    11          TRUE
```

For rounds 7 and 8 the information is published separately, as sample design data
files:

```{r}
ess_data_files(kind = "sample_design")[, .(round, file_label, doi, n_variables)]
#>    round    file_label                    doi n_variables
#>    <int>        <char>                 <char>       <int>
#> 1:     7  ESS7SDDFe1_2  10.21338/ess7sddfe1_2          10
#> 2:     8 ESS8SDDFe01_1 10.21338/ess8sddfe01_1          10
```

Two files, for two rounds. **Rounds 1–6 have neither**: no `psu` or `stratum` in
the integrated file, and no SDDF published through this API. For those rounds you
can weight, and you cannot compute a design-based variance from what the API
serves. The honest options are to say so, to restrict the design-based part of
the analysis to rounds 7–11, or to look for the older SDDF releases on the ESS
web portal. Silently reporting simple-random-sample intervals for rounds 1–6 is
not among them.

## Rounds 9–11: the design is already there

```{r}
cols <- c("essround", "cntry", "idno", "anweight", "pspwght", "pweight",
          "psu", "stratum", "prob", "domain",
          "health", "eisced", "agea", "gndr")

dt <- ess_round(11, select = cols)
```

Two things to check before building a design object.

**Are the identifiers unique across countries?** `stratum` and `psu` are numbered
within each country's own sample frame, so the same integer means different
places in different countries. Combining rounds adds a second collision.

```{r}
dt[, .(strata = uniqueN(stratum), psus = uniqueN(psu)), by = cntry][1:5]

# Distinct values overall versus summed within country — equal only if no reuse.
dt[, uniqueN(stratum)]
dt[, uniqueN(stratum), by = cntry][, sum(V1)]
```

If those two numbers differ, the codes are reused and must be made unique. Doing
it unconditionally is harmless and safer than testing:

```{r}
dt[, stratum_id := paste(essround, cntry, stratum, sep = "_")]
dt[, psu_id := paste(essround, cntry, psu, sep = "_")]
```

**How much is missing?** `svydesign()` will not accept `NA` in the strata or
cluster variables:

```{r}
dt[, .(no_psu = sum(is.na(psu)), no_stratum = sum(is.na(stratum)),
       no_weight = sum(is.na(anweight))), by = cntry][no_psu > 0 | no_stratum > 0]
```

Countries with no design information cannot join a clustered design. Drop them
from the design-based part and name them in the write-up.

### One country

Start here, because a single country's design is unambiguous:

```{r}
options(survey.lonely.psu = "adjust")

nl <- dt[cntry == "NL" & !is.na(psu) & !is.na(stratum) & !is.na(anweight)]

des_nl <- svydesign(
  ids = ~psu_id,
  strata = ~stratum_id,
  weights = ~anweight,
  data = nl,
  nest = TRUE
)

svymean(~I(health >= 4), des_nl, na.rm = TRUE)
confint(svymean(~I(health >= 4), des_nl, na.rm = TRUE))
```

`nest = TRUE` tells **survey** that PSU codes are nested inside strata rather
than being unique in their own right. `survey.lonely.psu = "adjust"` handles
strata that contain a single PSU, which occur in ESS samples and would otherwise
make the variance undefined; `"adjust"` centres such strata on the grand mean.
It is the conventional choice, and it is a choice — record it.

Compare against the naive interval to see the size of the correction:

```{r}
naive <- nl[!is.na(health), .(p = mean(health >= 4), n = .N)]
naive[, se := sqrt(p * (1 - p) / n)]
naive[, .(p, lower = p - 1.96 * se, upper = p + 1.96 * se)]
```

The design effect makes it concrete — how many independent observations this
clustered sample is worth:

```{r}
svymean(~I(health >= 4), des_nl, na.rm = TRUE, deff = TRUE)
```

### All countries at once

For cross-national estimates, country is a stratifying dimension: each country
ran its own sample, so strata nest inside countries and no PSU spans a border.
The combined identifiers built above already encode that.

Derive every analysis variable *before* building the design. `svydesign()` takes a
copy of the data, so a column added to `usable` afterwards is not visible to the
design object.

```{r}
usable <- dt[!is.na(psu) & !is.na(stratum) & !is.na(anweight)]

usable[eisced %in% c(0L, 55L), eisced := NA_integer_]
usable[, edu3 := cut(eisced, c(0, 2, 4, 7), labels = c("Low", "Medium", "High"))]

des <- svydesign(
  ids = ~psu_id,
  strata = ~stratum_id,
  weights = ~anweight,
  data = usable,
  nest = TRUE
)

svyby(~I(health >= 4), ~cntry, des, svymean, na.rm = TRUE, vartype = c("se", "ci"))
```

That is the table from `vignette("health-inequality")`, now with intervals you can
defend. And the regression from that vignette, done properly:

```{r}
fit <- svyglm(I(health >= 4) ~ edu3 + agea + factor(gndr) + cntry,
              design = subset(des, !is.na(edu3)),
              family = quasibinomial())

summary(fit)
exp(cbind(OR = coef(fit), confint(fit)))
```

`svyglm()` replaces both the point estimates and the variance, so the intervals
above account for weighting, stratification and clustering together. Note the
`subset(des, ...)` rather than filtering the data frame first: subsetting the
design keeps the full stratum structure for the variance calculation, whereas
dropping rows from the data and rebuilding loses strata that emptied.

## Rounds 7 and 8: merge the SDDF file

The SDDF file holds the design variables and the keys to attach them:

```{r}
sddf_files <- ess_data_files(kind = "sample_design")
sddf7 <- sddf_files[round == 7]

ess_variables(datafile_id = sddf7$datafile_id,
              datafile_version = sddf7$datafile_version)[, .(variable_name, variable_label)]
#>     variable_name                     variable_label
#>            <char>                             <char>
#>  1:          name                   Title of dataset
#>  2:      essround                          ESS round
#>  3:       edition                            Edition
#>  4:      proddate                    Production date
#>  5:         cntry                            Country
#>  6:          idno Respondent's identification number
#>  7:           psu              Primary sampling unit
#>  8:        domain                             Domain
#>  9:       stratum                   Sampling stratum
#> 10:          prob               Sampling probability
```

`cntry` and `idno` together identify a respondent, which is the join key.
`ess_data()` fetches the file by DOI, since `ess_round()` covers only the file
kinds that are round-level questionnaires:

```{r}
sddf <- ess_data(sddf7$doi, select = c("cntry", "idno", "psu", "stratum", "prob"))

main7 <- ess_round(7, select = c("essround", "cntry", "idno", "anweight",
                                 "health", "eisced", "agea", "gndr"))
```

Check the join before trusting it. `idno` is unique within a country, not across
the survey, so the key is the pair:

```{r}
sddf[, .N, by = .(cntry, idno)][N > 1L]        # expect no rows
main7[, .N, by = .(cntry, idno)][N > 1L]       # expect no rows

setdiff(unique(main7$cntry), unique(sddf$cntry))
setdiff(unique(sddf$cntry), unique(main7$cntry))
```

Countries in the integrated file but not in the SDDF have no design data for that
round — the same situation as the `NA`s in round 11, arriving by a different
route.

```{r}
merged <- merge(main7, sddf, by = c("cntry", "idno"), all.x = TRUE)

merged[, .(n = .N, matched = sum(!is.na(psu))), by = cntry][matched < n]
```

`all.x = TRUE` keeps every respondent so that the unmatched ones stay visible and
countable. An inner join would drop them and quietly shrink the sample.

From here it is the round 11 procedure:

```{r}
merged[, stratum_id := paste(cntry, stratum, sep = "_")]
merged[, psu_id := paste(cntry, psu, sep = "_")]

des7 <- svydesign(
  ids = ~psu_id, strata = ~stratum_id, weights = ~anweight,
  data = merged[!is.na(psu) & !is.na(stratum)], nest = TRUE
)

svyby(~I(health >= 4), ~cntry, des7, svymean, na.rm = TRUE, vartype = "ci")
```

Note that `anweight` comes from the integrated file, not from the SDDF. `prob`,
the sampling probability, is the raw inclusion probability; the ESS's published
weights already incorporate it along with non-response and post-stratification
adjustments. Use the published weight, not `1 / prob`.

## Comparing two rounds

For the round 7 versus round 11 comparison, build one design over the stacked
data with the round in the identifiers — which is what the `essround` prefix
earlier was for:

```{r}
pooled <- rbindlist(list(
  merged[, .(essround, cntry, idno, anweight, psu, stratum, health, eisced, agea, gndr)],
  dt[, .(essround, cntry, idno, anweight, psu, stratum, health, eisced, agea, gndr)]
), use.names = TRUE)

pooled[, stratum_id := paste(essround, cntry, stratum, sep = "_")]
pooled[, psu_id := paste(essround, cntry, psu, sep = "_")]

pooled[eisced %in% c(0L, 55L), eisced := NA_integer_]
pooled[, edu3 := cut(eisced, c(0, 2, 4, 7), labels = c("Low", "Medium", "High"))]
pooled[, round_f := factor(essround)]

des_both <- svydesign(
  ids = ~psu_id, strata = ~stratum_id, weights = ~anweight,
  data = pooled[!is.na(psu) & !is.na(stratum)], nest = TRUE
)

svyby(~I(health >= 4), ~essround, des_both, svymean, na.rm = TRUE, vartype = "ci")

fit_both <- svyglm(I(health >= 4) ~ round_f * edu3 + agea + cntry,
                   design = subset(des_both, !is.na(edu3)),
                   family = quasibinomial())

summary(fit_both)
```

The interaction is the test the user story asked for: whether the educational
gradient differs between 2014 and 2023. Two cross-sections nine years apart
treated as independent samples — which they are, the ESS being repeated
cross-sectional rather than a panel.

## What this vignette worked around

* Design variables are in the integrated file from round 9 only; rounds 7 and 8
  publish them as separate SDDF files; rounds 1–6 have neither through this API.
* `psu` and `stratum` are numbered within country, so codes collide when
  countries or rounds are pooled. Build compound identifiers.
* `nest = TRUE` and `survey.lonely.psu = "adjust"` are both needed in practice
  with ESS samples.
* The SDDF join key is `cntry` + `idno` together; `idno` alone is not unique.
  Use `all.x = TRUE` so unmatched respondents stay countable.
* Use the published `anweight`, not `1 / prob`.
* `subset()` the design rather than filtering the data and rebuilding, or the
  variance loses emptied strata.
