---
title: "Getting data from the European Social Survey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting data from the European Social Survey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # Every chunk below talks to the live ESS API, so none of them run at build
  # time. The output shown is from a real session.
  eval = FALSE
)
```

This vignette walks through configuring the package, working out what data
exists, downloading it, and handling the parts of the ESS that need a little
care: missing values, value labels, and country selection.

```{r setup}
library(essurvey2)
library(data.table)
```

## Configuring your user ID

Every request for data carries a user ID, which the ESS uses to record download
statistics. It is not a password and it does not authenticate you, but the API
refuses requests without one.

Get yours by logging in at <https://ess.sikt.no/en/api>. Then put it in your user
`.Renviron` file, which `usethis::edit_r_environ()` opens:

```
ESS_USER_ID=your-user-id-here
```

Restart R afterwards — `.Renviron` is read only at startup. A system environment
variable of the same name works identically, and is the better choice on a shared
or containerised machine.

For a single session, or to try the package out before deciding where to put the
ID:

```{r}
ess_set_user_id("your-user-id-here")
```

`ess_config()` shows what the package resolved, with the ID masked so the output
is safe to paste into a bug report:

```{r}
ess_config()
#> ── essurvey2 configuration
#> data API:  <https://api.ess.sikt.no>
#> catalogue: <https://api.nsd.no/graphql>
#> user ID:   "a1b2c3d4****************************" (from ESS_USER_ID)
#> cache:     'C:/Users/you/AppData/Local/R/cache/R/essurvey2'
```

## Looking before you download

An ESS integrated file is a few hundred thousand rows by seven hundred columns,
so it pays to know what you are asking for. Everything in this section reads the
metadata catalogue, which needs no user ID and transfers almost nothing.

Start with the rounds:

```{r}
ess_rounds()[, .(round, file_label, edition, description)]
#>     round file_label edition                                       description
#>  1:     1  ESS1e06_7     6.7 ESS round 1 - 2002. Immigration, Citizen involve…
#>  2:     2  ESS2e03_6     3.6 ESS round 2 - 2004. Health and care, Economic mo…
#> ...
#> 11:    11 ESS11e04_2     4.2 ESS round 11 - 2023. Social inequalities in heal…
```

Note the `edition` column. Editions are revised as the ESS corrects and extends
its data, and each edition has its own DOI. This package always resolves the
current one from the catalogue, which is why no DOI is hard-coded anywhere in it —
and why hard-coding one in your own script is a bad idea.

Countries, with the codes that appear in the data's `cntry` column:

```{r}
ess_countries()[1:5]
#>    country_code country_name n_rounds          rounds
#> 1:           AL      Albania        1               6
#> 2:           AT      Austria       10 1,2,3,4,5,6,...
#> ...

ess_country_rounds("NL")[, .(round, file_label, doi)]
```

Themes tell you which rounds asked about what. Six are core themes, repeated
every round; the rest are rotating modules:

```{r}
ess_themes()[is_core == FALSE, .(theme_name)]
ess_theme_rounds("Justice")[, .(round, description)]
```

### Finding variables

`ess_variables()` lists a data file's variables with their labels, which is how
you find the names to pass to `select`:

```{r}
vars <- ess_variables(11)
nrow(vars)
#> [1] 709

vars[grepl("trust", variable_label, ignore.case = TRUE),
     .(variable_name, variable_label)]
#>    variable_name                                         variable_label
#> 1:       ppltrst Most people can be trusted or you can't be too careful
#> 2:       trstprl                          Trust in country's parliament
#> 3:       trstlgl                              Trust in the legal system
#> ...
```

For one variable, `ess_variable_info()` gives the question as it was put to
respondents, and what each code means:

```{r}
info <- ess_variable_info("ppltrst", round = 11)
info$question
#> [1] "Using this card, generally speaking, would you say that most people can
#>      be trusted, or that you can't be too careful in dealing with people?"

info$codes
#>     value                      label is_missing
#>  1:     0   You can't be too careful      FALSE
#>  2:     1                          1      FALSE
#> ...
#> 12:    77                    Refusal       TRUE
#> 13:    88                 Don't know       TRUE
#> 14:    99                  No answer       TRUE
```

## Downloading a round

```{r}
dt <- ess_round(11, select = c("essround", "cntry", "idno", "agea", "gndr", "ppltrst"))
#> ℹ Downloading "10.21338/ess11e04_2" as parquet, missing values recoded.
#> ✔ Downloaded 12.4 MB to '…/essurvey2/10.21338_ess11e04_2__recoded.parquet'.

dim(dt)
#> [1] 50116     6
class(dt)
#> [1] "data.table" "data.frame"
```

`select` is worth using. Parquet is columnar, so only the requested columns are
decoded — reading six of them takes a fraction of the time and memory that all
709 would.

Several rounds come back stacked, distinguished by `essround`:

```{r}
trend <- ess_round(9:11, select = c("essround", "cntry", "ppltrst", "pspwght"))

trend[, .(trust = round(weighted.mean(ppltrst, pspwght, na.rm = TRUE), 2)),
      by = essround][order(essround)]
```

Rounds are stacked with `fill = TRUE`, so a variable that does not exist in one
of them is `NA` for that round's rows rather than an error.

`ess_all_rounds()` fetches the lot. With `select` it is quite manageable; without
it, expect a few hundred megabytes.

```{r}
everything <- ess_all_rounds(select = c("essround", "cntry", "ppltrst", "pspwght"))
```

### Caching

The download above went to a cache directory. Asking again — in this session or
next month — reads from disk:

```{r}
dt <- ess_round(11, select = c("essround", "cntry"))
#> ℹ Using cached "10.21338/ess11e04_2" (12.4 MB).

ess_cache_list()[, .(file, size)]
#>                                    file    size
#> 1: 10.21338_ess11e04_2__recoded.parquet 12.4 MB
```

The cache key covers the DOI, the format and the missing-value treatment, so
`recode_missings = TRUE` and `FALSE` are separate entries rather than one
overwriting the other.

```{r}
options(essurvey2.cache_dir = "D:/data/ess-cache")  # put it somewhere else
options(essurvey2.cache_dir = FALSE)                # or switch it off
ess_cache_clear()                                   # or empty it
```

## Missing values

The ESS codes non-responses as out-of-range values, and the codes differ from
variable to variable: `ppltrst` uses 77, 88 and 99, while `agea` uses 999. This
is a trap for anyone who assumes a single convention.

By default `essurvey2` asks the API to convert them, so they arrive as `NA`:

```{r}
dt <- ess_round(11, select = c("essround", "agea", "ppltrst"))
max(dt$agea, na.rm = TRUE)
#> [1] 90
sum(is.na(dt$agea))
#> [1] 393
```

Compare that with the original codes:

```{r}
raw <- ess_round(11, select = c("essround", "agea", "ppltrst"),
                 recode_missings = FALSE)
max(raw$agea)
#> [1] 999
sum(is.na(raw$agea))
#> [1] 0
```

One quirk is worth knowing about. The API treats `recodeMissingValues` as a flag
rather than a value, so *any* value switches recoding on —
`recodeMissingValues=false` still recodes. The only way to turn it off is to omit
the parameter, which is what `recode_missings = FALSE` does.

To do the recoding locally instead — because you want the codes for some
variables and `NA` for others, or want to see which codes were affected — use
`ess_recode_missings()`. It reads each variable's own missing codes from the
catalogue:

```{r}
ess_missing_codes(11, variables = c("agea", "gndr", "ppltrst"))
#>    variable_name value         label
#> 1:       ppltrst    77       Refusal
#> 2:       ppltrst    88    Don't know
#> 3:       ppltrst    99     No answer
#> 4:          gndr     9     No answer
#> 5:          agea   999 Not available

ess_recode_missings(raw)
#> ✔ Recoded 537 values to `NA` across 2 columns.
```

Because this follows `data.table` conventions, `raw` is modified **in place**.
Pass `copy(raw)` if you want to keep the original. Note also that it recoded only
two columns: `gndr`'s missing code, 9, does not actually occur in round 11.

## Value labels

Parquet files carry codes, not labels — a `gndr` of 1, not "Male". That differs
from the earlier `essurvey` package, which read SPSS files with their labels
attached. The labels live in the catalogue:

```{r}
ess_value_labels(11, variables = "gndr")
#>    variable_name value     label is_missing
#> 1:          gndr     1      Male      FALSE
#> 2:          gndr     2    Female      FALSE
#> 3:          gndr     9 No answer       TRUE
```

`ess_as_factor()` applies them, again in place:

```{r}
dt <- ess_round(11, select = c("essround", "cntry", "gndr"))
ess_as_factor(dt, variables = "gndr")
#> ✔ Converted 1 column to factor.

dt[, .N, by = gndr]
#>      gndr     N
#> 1:   Male 23089
#> 2: Female 27027
```

Only variables with a code list are converted, and codes flagged as missing
become `NA` rather than a level — pass `keep_missing = TRUE` to keep them.
Variables with more than `max_levels` codes are left alone, so a near-continuous
variable does not silently become a factor with hundreds of levels.

Think before converting a scale such as `ppltrst`: as a factor you can no longer
take its mean.

## Selecting a country

`ess_country()` returns one country's respondents:

```{r}
nl <- ess_country("NL", rounds = 10:11, select = c("essround", "agea", "ppltrst"))
#> ℹ Fetching 2 data files: "ESS10e03_3" and "ESS11e04_2".
#> ✔ 3165 respondents from "NL" across 2 rounds.

nl[, .(n = .N, trust = round(mean(ppltrst, na.rm = TRUE), 2)), by = essround]
#>    essround    n trust
#> 1:       10 1470  6.54
#> 2:       11 1695  6.26
```

Be aware of what this does. The current API publishes integrated files only —
there are no per-country downloads any more — so each round is fetched in full
and then filtered on `cntry`. The cache means you pay for that once, and `select`
keeps it small, but the first call transfers the whole round.

Omitting `rounds` uses every round the country took part in, as reported by
`ess_country_rounds()`. Asking for a round a country skipped warns rather than
returning zero rows silently:

```{r}
ess_country("AL", rounds = 11, select = "essround")
#> Warning: "AL" did not take part in ESS round 11.
#> ℹ That round will contribute no rows.
```

## The other data files

Each round publishes more than the integrated file. `ess_data_files()` shows what
a round has, and `ess_file_kinds()` explains the names:

```{r}
ess_data_files(11)[, .(file_kind, file_label, doi, n_variables)]
#>        file_kind    file_label                    doi n_variables
#> 1:       alcohol ESS11ALCe02_1 10.21338/ess11alce02_1         753
#> 2: contact_forms  ESS11CFe04_1  10.21338/ess11cfe04_1         247
#> 3:    integrated    ESS11e04_2    10.21338/ess11e04_2         709
#> 4:   interviewer ESS11INTe04_1 10.21338/ess11inte04_1          25
```

Round 10 introduced self-completion modes, which have their own integrated file:

```{r}
sc <- ess_round(10, kind = "self_completion", select = c("essround", "cntry"))
```

Anything not covered by `ess_round()` can be fetched straight by DOI:

```{r}
cf <- ess_data("10.21338/ess11cfe04_1")
```

Two kinds deserve a mention. **Sample design data** holds the design weights and
clustering variables needed for correct standard errors — what the old package's
`import_sddf_country()` fetched. Only some rounds publish it:

```{r}
ess_data_files(kind = "sample_design")[, .(round, file_label, doi)]
#>    round    file_label                    doi
#> 1:     7  ESS7SDDFe1_2  10.21338/ess7sddfe1_2
#> 2:     8 ESS8SDDFe01_1 10.21338/ess8sddfe01_1

sddf <- ess_data("10.21338/ess7sddfe1_2")
names(sddf)
#> [1] "name" "essround" "edition" "proddate" "cntry" "idno" "psu" "domain"
#> [9] "stratum" "prob"
```

And some countries are not in an integrated file at all — because their fieldwork
ran outside the round's window, or because they lack design weights — and are
published as separate files:

```{r}
ess_data_files(kind = "country")[, .(round, file_country, description)]
#>    round file_country                                             description
#> 1:     2           IT       ESS2 - Italy country file from main questionnaire
#> 2:     3           LV                       ESS3 - Latvia (no design weights)
#> 3:     3           RO                      ESS3 - Romania (no design weights)
#> 4:     4           AT  ESS4 - Austria (Fieldwork period 01.11.10 to 28.02.11)
#> 5:     4           LT                    ESS4 - Lithuania (no design weights)
#> 6:     5           AT ESS5 - Austria (fieldwork period 24.05.13 to 10.10.13)
#> 7:     9           RO            ESS9 - Romania (participating on pilot basis)
#> 8:     9           AL              ESS9 - Albania (interview quality issues)

at5 <- ess_data("10.21338/ess5ate1_1")
```

This is worth knowing before concluding that a country is missing from a round.
`ess_country("AT")` will not return round 5 Austria, because that data is not in
the round 5 integrated file — it is in `ESS5ATe1_1`. The descriptions above say
why each one was held out, and whether it is safe to pool with the rest is a
judgement for you rather than this package.

### When the catalogue contradicts itself

A DOI should identify one file. In round 8 the catalogue currently gives the same
DOI to the contact forms and to the interview time data, so `ess_data_files()`
warns:

```{r}
ess_data_files(8)
#> Warning: The ESS catalogue gives 2 data files the same DOI "10.21338/ess8timee01".
#> ✖ "ESS8TIMEe01" and "ESS8CFe03"
#> ℹ Downloading it returns one file, which may not be the one you asked for.
```

Downloading that DOI returns whichever file it really points at. This package
reports the inconsistency rather than guessing, since guessing wrong would hand
you the wrong data silently.

## Other formats

Parquet is the default and the fastest, but the API also serves CSV, SPSS and
Stata. `sav` and `dta` are read with **haven**, which must be installed.

```{r}
csv <- ess_data("10.21338/ess11inte04_1", format = "csv")

# Keep the SPSS file rather than reading it.
path <- ess_download_file("10.21338/ess11e04_2", path = "ess11.sav", format = "sav")
```

## When something goes wrong

Failures are raised as classed conditions, so they can be caught specifically:

| Class | Meaning |
|---|---|
| `essurvey2_error_user_id` | no user ID configured, or the API rejected it |
| `essurvey2_error_doi` | no such data file, or a study DOI was passed |
| `essurvey2_error_server` | server-side failure; large files are also refused here |
| `essurvey2_error_catalogue` | the metadata catalogue returned an error |
| `essurvey2_error_offline` | the catalogue could not be reached |

```{r}
tryCatch(
  ess_data("10.21338/NSD-ESS10-2020"),
  essurvey2_error_doi = function(e) message("Not a data file DOI")
)
```

Messages from the API include a request ID, which is what ESS support will ask
for.

Two things to know about the endpoint. It only accepts **data file** DOIs, not
study-level ones such as `10.21338/NSD-ESS10-2020` — `ess_data_files()` returns
the right kind. And it cannot serve the very largest files at all; those still
have to come from the web portal.

## Going further

The metadata catalogue holds considerably more than this package wraps. `ess_gql()`
sends arbitrary GraphQL queries, and the schema is introspectable:

```{r}
ess_gql("
  query($id: ID!, $version: Int) {
    search {
      seriesMetadata(id: $id, version: $version, instance: PUBLISHED, agencyId: INT_ESSERIC) {
        citation { abstract { en } }
      }
    }
  }",
  variables = ess_series_id()
)
```

`ess_series()` lists the other series published by the ESS ERIC — the CRONOS web
panels, ESS Multilevel Data, ICOS Cities — whose IDs can be used the same way.

Catalogue answers are memoised for the session. `ess_catalogue_refresh()` forgets
them, which is what you want if a new round is published while your session is
open.
