---
title: "Building one comparable variable across rounds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building one comparable variable across rounds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

> **User story.** As a survey methodologist preparing a pooled ESS file, I need
> to know which variant of each variable exists in each round and how its codes
> are defined, so that I can build one comparable measure instead of silently
> stacking incompatible classifications.

Catalogue output below is real, from a session against the live catalogue.
Anything computed from survey data is shown as code only — see
`vignette("trust-trends")` for why numbers are not pasted in.

```{r setup}
library(essurvey2)
library(data.table)
```

## The failure this avoids

`ess_round(1:11, select = c("essround", "cntry", "edulvla"))` does not error. It
returns a `data.table` in which `edulvla` is populated for rounds 1 to 4 and `NA`
for rounds 5 to 11, because `ess_round()` stacks with `fill = TRUE`. Regress
something on it and you have fitted a model to the first four rounds while
believing you used eleven.

The ESS renames and reclassifies variables between rounds — that is a
consequence of harmonising 39 countries' education systems, ballot papers and
income brackets, not carelessness. What the analyst owes the data is a check
before the download, and `ess_variables()` is cheap enough to run eleven times.

## An availability matrix

One call per round, no survey data, no user ID:

```{r}
candidates <- c("edulvla", "edulvlb", "eisced", "eduyrs",
                "hinctnta", "hincfel",
                "wrkprty", "pbldmn", "sgnptit")

availability <- rbindlist(lapply(1:11, function(r) {
  present <- ess_variables(r)$variable_name
  data.table(round = r, variable_name = candidates,
             present = candidates %in% present)
}))

dcast(availability, variable_name ~ round, value.var = "present")
#> Key: <variable_name>
#>    variable_name     1     2     3     4     5     6     7     8     9    10    11
#> 1:       edulvla  TRUE  TRUE  TRUE  TRUE FALSE FALSE FALSE FALSE FALSE FALSE FALSE
#> 2:       edulvlb FALSE FALSE FALSE FALSE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE
#> 3:        eduyrs  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE
#> 4:        eisced  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE
#> 5:       hincfel  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE
#> 6:      hinctnta FALSE FALSE FALSE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE
#> 7:        pbldmn  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE FALSE FALSE
#> 8:       sgnptit  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE
#> 9:       wrkprty  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE FALSE FALSE
```

Four distinct patterns, each needing a different decision:

* **Replaced at a round boundary.** `edulvla` (rounds 1–4) became `edulvlb`
  (rounds 5–11). They are not the same classification; concatenating the columns
  would be wrong.
* **Available throughout.** `eisced` and `eduyrs` span the series. That is the
  answer to the education problem, with one catch — below.
* **Introduced mid-series.** `hinctnta` starts at round 4. Rounds 1–3 measured
  income differently, so a decile series simply starts in 2008.
* **Discontinued.** `wrkprty` and `pbldmn` stop after round 9. A political
  participation index built on them ends there; `sgnptit` continues.

That first pattern is exactly the trap in the opening paragraph, and the matrix
is what turns it into a decision you made on purpose.

## Education: use the harmonised variable, then look at its codes

`eisced` is the ESS's own harmonisation of education into a common seven-level
scheme, and the matrix above shows it in all eleven rounds. Before using it as a
gradient, read the code list:

```{r}
ess_variable_info("eisced", round = 11)$codes
#>      value                                               label is_missing
#>     <char>                                              <char>     <lgcl>
#>  1:      0             Not possible to harmonise into ES-ISCED      FALSE
#>  2:      1              ES-ISCED I , less than lower secondary      FALSE
#>  3:      2                        ES-ISCED II, lower secondary      FALSE
#>  4:      3           ES-ISCED IIIb, lower tier upper secondary      FALSE
#>  5:      4           ES-ISCED IIIa, upper tier upper secondary      FALSE
#>  6:      5        ES-ISCED IV, advanced vocational, sub-degree      FALSE
#>  7:      6     ES-ISCED V1, lower tertiary education, BA level      FALSE
#>  8:      7 ES-ISCED V2, higher tertiary education, >= MA level      FALSE
#>  9:     55                                               Other      FALSE
#> 10:     77                                             Refusal       TRUE
#> 11:     88                                          Don't know       TRUE
#> 12:     99                                           No answer       TRUE
```

Note the `is_missing` column. Codes 77, 88 and 99 are flagged, so
`recode_missings = TRUE` — the default — turns them into `NA`. Codes **0** and
**55** are *not* flagged, and will not be touched by any missing-value handling,
by this package or by the API.

They are still meaningless as levels of education. "Not possible to harmonise
into ES-ISCED" is a documentation outcome; "Other" is a residual. Leave them in
and a mean of `eisced` is arithmetic over a scale that runs 0, 1…7, 55 — the 55s
alone will dominate it, and `max(eisced)` will report 55 as the highest level of
education in Europe.

```{r}
dt <- ess_round(11, select = c("essround", "cntry", "eisced", "agea", "anweight"))

dt[, .N, by = eisced][order(eisced)]     # look for 0 and 55 before anything else

dt[eisced %in% c(0L, 55L), eisced := NA_integer_]
```

This is the general point, not a fact about `eisced`. `is_missing` marks
*non-response*. It does not mark codes that are substantively unusable for your
particular analysis, and no automatic step can, because that depends on the
analysis. Read the code list.

## Ordered factors from the catalogue

With the unusable codes gone, `ess_as_factor()` labels the rest. It reads the
labels from the catalogue, since parquet carries codes only:

```{r}
ess_as_factor(dt, variables = "eisced")
#> ✔ Converted 1 column to factor.

dt[, .N, by = eisced]
```

`ess_as_factor()` modifies `dt` in place, `data.table` fashion. It also drops
missing-flagged codes rather than making them levels, and leaves alone any column
with more than `max_levels` codes — so a near-continuous variable does not become
a factor with hundreds of levels. Columns that are already factors are left alone
too, so re-running this chunk converts each column once rather than blanking it.

Keep a numeric copy if you want both a gradient and a labelled table:

```{r}
dt[, eisced_num := as.integer(eisced)]   # 1..8 over the surviving levels
```

For a linear gradient, `eduyrs` — years of full-time education, present in all
eleven rounds — is often the easier instrument, at the cost of comparability
between school systems.

## Country-specific variables

Some questions cannot be harmonised at all, because the response options are a
national ballot paper. The catalogue lists both the template and every country's
realisation of it:

```{r}
vars <- ess_variables(11)
vars[grepl("^prtv", variable_name), .(variable_name, variable_label)][1:15]
#>     variable_name                                                          variable_label
#>            <char>                                                                  <char>
#>  1:       prtvtxx                                  Party voted for in last national election, [country]
#>  2:      prtvtdat                                    Party voted for in last national election, Austria
#>  3:      prtvtebe                                    Party voted for in last national election, Belgium
#>  4:      prtvtfbg                                   Party voted for in last national election, Bulgaria
#> ...
#>  9:      prtvtffr                          Party voted for in last national election, France (ballot 1)
#> 10:      prtvgde1                                  Party voted for in last national election 1, Germany
#> 11:      prtvgde2                                  Party voted for in last national election 2, Germany
```

`prtvtxx` is a template, not a column: the `xx` stands for a country suffix that
also encodes which version of the question was asked. Passing `prtvtxx` to
`select` is an error naming the column, because `select` is checked against the
file itself rather than against the catalogue's list.

The suffix letter changes between rounds when a country's party list changes, so
the name is not stable across the series either. Read the columns you need per
round and per country:

```{r}
party_cols <- vars[grepl("^prtv", variable_name) &
                     variable_name != "prtvtxx", variable_name]

# Germany has two columns in round 11 — first and second vote.
vars[variable_name %in% c("prtvgde1", "prtvgde2"), .(variable_name, variable_label)]
```

`ess_variable_info()` gives each one its own party list, which is the mapping you
would need for any cross-national recoding — into left/right blocs, government
versus opposition, or whatever the analysis calls for:

```{r}
ess_variable_info("prtvtdat", round = 11)$codes
```

That recoding is a research decision this package deliberately does not make for
you.

## The variable list is a superset

`ess_variables()` returns the catalogue's documentation of a data file, which
includes entries with no column in that file: the `xx` templates above, and
questionnaire items such as the higher household-grid slots that exist in the
instrument but were never filled.

So the matrix earlier tells you what the catalogue documents, which is a
necessary check but not a sufficient one. `select` is validated against the
parquet file, and that is the authority:

```{r}
tryCatch(
  ess_round(11, select = c("essround", "prtvtxx")),
  essurvey2_error_select = function(e) conditionMessage(e)
)
```

Belt and braces for a scripted pipeline — resolve the columns against the file
once, then read:

```{r}
have <- names(ess_read_file(ess_download_file(
  ess_data_files(11, kind = "integrated")$doi
)))

setdiff(candidates, have)
```

## A pooled file, assembled deliberately

Putting it together: one variable set, checked per round, with the rounds that
cannot supply it excluded rather than filled with `NA`.

```{r}
wanted <- c("essround", "cntry", "idno", "anweight", "eisced", "eduyrs",
            "agea", "gndr", "hinctnta", "ppltrst")

usable <- rbindlist(lapply(1:11, function(r) {
  present <- ess_variables(r)$variable_name
  data.table(round = r, missing = paste(setdiff(wanted, present), collapse = ", "))
}))
usable
#>    round  missing
#> 1:     1 hinctnta
#> 2:     2 hinctnta
#> 3:     3 hinctnta
#> 4:     4
#> ...

pooled <- ess_round(4:11, select = wanted)

pooled[eisced %in% c(0L, 55L), eisced := NA_integer_]
pooled[, .N, by = essround]
```

Rounds 1–3 are excluded because they cannot supply `hinctnta`, and the table
above is the record of that decision. Whether to run the analysis twice — once
on rounds 4–11 with income, once on 1–11 without — is a judgement about your
question, not about the data.

## What this vignette worked around

* `edulvla` (rounds 1–4) and `edulvlb` (rounds 5–11) are different
  classifications of the same concept. `eisced` spans the series; use it.
* `eisced` codes 0 and 55 are not flagged as missing but are unusable as
  education levels. `is_missing` means non-response, nothing more.
* `hinctnta` begins at round 4; `wrkprty` and `pbldmn` end at round 9.
* `ess_variables()` is a superset of a file's columns. `prtvtxx` is a template.
  `select` is checked against the file, and is the authority.
* `ess_round()` stacks with `fill = TRUE`, so an absent variable arrives as `NA`
  rather than as an error. Check before you download.
