---
title: "The educational gradient in self-rated health"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The educational gradient in self-rated health}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

> **User story.** As a social epidemiologist, I need self-rated health broken
> down by education and household income for the two ESS rounds that carried the
> health-inequality module, so that I can test whether the educational gradient
> in poor health widened between 2014 and 2023.

Catalogue output below is real. Results from survey data are shown as code only —
see `vignette("trust-trends")` for why.

```{r setup}
library(essurvey2)
library(data.table)
```

## Which rounds carry the module

Health inequality is a rotating module, so the first question is when it ran:

```{r}
ess_theme_rounds("Social inequalities in health")[, .(round, description)]
#>    round                                                                       description
#>    <int>                                                                            <char>
#> 1:     7                    ESS round 7 - 2014. Immigration, Social inequalities in health
#> 2:    11 ESS round 11 - 2023. Social inequalities in health, Gender in contemporary Europe
```

Rounds 7 (2014) and 11 (2023): a nine-year gap, one repeat of the module, and
therefore a comparison of two cross-sections rather than a trend. Say so in the
write-up — two points do not make a trajectory.

There is a separate, older health theme, which is worth not confusing with this
one:

```{r}
ess_theme_rounds("Health and care")[, .(round, description)]
#>    round                                                                       description
#>    <int>                                                                            <char>
#> 1:     2 ESS round 2 - 2004. Health and care, Economic morality, Family work and wellbeing
```

Round 2's module asked different questions. The outcome used below — `health` —
is a core variable present in every round, but the module's supporting variables
are not, which is the reason to stay inside rounds 7 and 11.

## What the module contains

Variables come grouped, and the group is the module:

```{r}
v11 <- ess_variables(11)
v11[group_name == "healthineq", .N]
#> [1] 113

v11[group_name == "healthineq", .(variable_name, variable_label)][1:5]
#>    variable_name                                                variable_label
#>           <char>                                                        <char>
#> 1:       ctrlife                How much control over life in general nowadays
#> 2:       etfruit                 How often eat fruit, excluding drinking juice
#> 3:        eatveg         How often eat vegetables or salad, excluding potatoes
#> 4:        dosprt Do sports or other physical activity, how many of last 7 days
#> 5:       cgtsmok                                   Cigarette smoking behaviour
```

Round 7 has the same group, and comparing the two lists is how you find the
module's stable core:

```{r}
v7 <- ess_variables(7)

both <- intersect(v7[group_name == "healthineq", variable_name],
                  v11[group_name == "healthineq", variable_name])
length(both)

setdiff(v11[group_name == "healthineq", variable_name],
        v7[group_name == "healthineq", variable_name])
```

Anything in that `setdiff` is round 11 only and cannot enter a comparison.

## Reading the outcome's codes before using it

```{r}
info <- ess_variable_info("health", round = 11)
info$question
#> [1] "How is your health in general? Would you say it is..."

info$codes
#>     value      label is_missing
#>    <char>     <char>     <lgcl>
#> 1:      1  Very good      FALSE
#> 2:      2       Good      FALSE
#> 3:      3       Fair      FALSE
#> 4:      4        Bad      FALSE
#> 5:      5   Very bad      FALSE
#> 6:      7    Refusal       TRUE
#> 7:      8 Don't know       TRUE
#> 8:      9  No answer       TRUE
```

**The scale runs from good to bad.** 1 is *Very good*; 5 is *Very bad*. A higher
number means worse health, which is the opposite of what "health" suggests and
the opposite of most ESS scales, where higher means more of the thing named. A
mean of `health` that goes up over time is health getting *worse*.

Either reverse it on the way in, or define a dichotomy whose direction is in its
name. The second is easier to report and easier to get right:

```{r}
# poor_health == TRUE means "Bad" or "Very bad".
```

The secondary outcome, activity limitation, runs the same way — and note that
"No" is 3, not 0:

```{r}
ess_variable_info("hlthhmp", round = 11)$codes
#>     value              label is_missing
#>    <char>             <char>     <lgcl>
#> 1:      1          Yes a lot      FALSE
#> 2:      2 Yes to some extent      FALSE
#> 3:                        No      FALSE
#> 4:      7            Refusal       TRUE
#> 5:      8         Don't know       TRUE
#> 6:      9          No answer       TRUE
```

## The stratifiers

Education, as `vignette("harmonising-variables")` sets out in more detail, needs
its unusable codes removed by hand:

```{r}
ess_variable_info("eisced", round = 11)$codes[value %in% c("0", "55", "77")]
#>     value                                   label is_missing
#>    <char>                                  <char>     <lgcl>
#> 1:      0 Not possible to harmonise into ES-ISCED      FALSE
#> 2:     55                                   Other      FALSE
#> 3:     77                                 Refusal       TRUE
```

Income is a decile, and the labels say something important about what kind:

```{r}
ess_variable_info("hinctnta", round = 11)$codes
#>      value           label is_missing
#>     <char>          <char>     <lgcl>
#>  1:      1  J - 1st decile      FALSE
#>  2:      2  R - 2nd decile      FALSE
#> ...
#> 10:     10 H - 10th decile      FALSE
#> 11:     77         Refusal       TRUE
#> 12:     88      Don't know       TRUE
#> 13:     99       No answer       TRUE
```

The bands behind those deciles are country-specific: respondents were shown a
card with their own country's income distribution. `hinctnta` is therefore a
household's *relative position within its own country*, not an amount. That makes
it directly comparable across countries — decile 1 means the same rank
everywhere — and useless as an absolute income. Do not deflate it, convert it, or
compare a Portuguese decile 8 with a Norwegian decile 8 in euros.

Item non-response on income is heavy in every survey, and the ESS is no
exception. Count it rather than dropping it silently:

```{r}
ess_missing_codes(11, variables = c("health", "eisced", "hinctnta", "agea", "gndr"))
#>    variable_name  value         label
#>           <char> <char>        <char>
#> 1:        health      7       Refusal
#> 2:        health      8    Don't know
#> 3:        health      9     No answer
#> 4:          gndr      9     No answer
#> 5:          agea    999 Not available
#> 6:        eisced     77       Refusal
#> 7:        eisced     88    Don't know
#> 8:        eisced     99     No answer
#> 9:      hinctnta     77       Refusal
```

Four different missing-code conventions in five variables: 7/8/9, 9 alone, 999,
and 77/88/99. This is why `recode_missings = TRUE` reads them per variable from
the catalogue instead of applying one rule.

## Downloading

```{r}
cols <- c("essround", "cntry", "idno", "anweight",
          "health", "hlthhmp", "eisced", "hinctnta", "agea", "gndr")

hi <- ess_round(c(7, 11), select = cols)
#> ℹ Fetching 2 data files: "ESS7e02_3" and "ESS11e04_2".

hi[, .N, by = essround]
```

`anweight` is the right weight for pooling countries; see `vignette("trust-trends")`
for the reasoning.

## Preparing the analysis variables

```{r}
hi[eisced %in% c(0L, 55L), eisced := NA_integer_]

hi[, poor_health := fifelse(health >= 4L, 1L, 0L, na = NA_integer_)]
hi[, limited := fifelse(hlthhmp %in% c(1L, 2L), 1L, 0L, na = NA_integer_)]

# Three bands are easier to report than eight levels, and less sparse per country.
hi[, edu3 := cut(eisced, breaks = c(0, 2, 4, 7), labels = c("Low", "Medium", "High"))]

hi[, income3 := cut(hinctnta, breaks = c(0, 3, 7, 10),
                    labels = c("Bottom 30%", "Middle 40%", "Top 30%"))]

hi[, age_band := cut(agea, breaks = c(14, 34, 49, 64, Inf),
                     labels = c("15-34", "35-49", "50-64", "65+"))]
```

`fifelse()` is used rather than `ifelse()` because it keeps the `NA` explicit:
`health >= 4L` is `NA` where `health` is `NA`, and `na = NA_integer_` says so
rather than quietly producing `0`.

Now account for the cases lost, before reporting anything:

```{r}
hi[, .(
  n = .N,
  no_health = sum(is.na(health)),
  no_edu = sum(is.na(eisced)),
  no_income = sum(is.na(hinctnta)),
  complete = sum(!is.na(health) & !is.na(eisced) & !is.na(hinctnta))
), by = essround]
```

If income non-response differs a lot between the two rounds, a change in the
gradient measured on complete cases may be a change in who answered. That
belongs in the write-up.

## The gradient

Weighted prevalence of poor health by education, per round, age-standardised
crudely by holding the age distribution out as a separate breakdown:

```{r}
gradient <- hi[!is.na(edu3) & !is.na(poor_health),
  .(n = .N, pct_poor = round(100 * weighted.mean(poor_health, anweight), 1)),
  by = .(essround, edu3)
][order(essround, edu3)]

dcast(gradient, edu3 ~ essround, value.var = "pct_poor")
```

The gradient's steepness is the quantity of interest — the low-education
prevalence minus the high-education prevalence:

```{r}
steepness <- dcast(gradient, essround ~ edu3, value.var = "pct_poor")
steepness[, gap := Low - High]
steepness
```

Age is the obvious confounder, since the less-educated are older on average in
most of these countries. Stratify:

```{r}
hi[!is.na(edu3) & !is.na(age_band) & !is.na(poor_health),
  .(pct_poor = round(100 * weighted.mean(poor_health, anweight), 1)),
  by = .(essround, age_band, edu3)
] |> dcast(age_band + edu3 ~ essround, value.var = "pct_poor")
```

And by country, for round 11, with a floor on cell size so that a gap computed
from a handful of respondents is not reported as though it were an estimate:

```{r}
by_country <- hi[essround == 11L & !is.na(edu3) & !is.na(poor_health),
  .(n = .N, pct_poor = 100 * weighted.mean(poor_health, anweight)),
  by = .(cntry, edu3)
]

wide <- dcast(by_country[n >= 50L], cntry ~ edu3, value.var = "pct_poor")
wide[, gap := round(Low - High, 1)]
wide[order(-gap)]
```

## Adjusting for income

Whether the educational gradient survives controlling for income is the
substantive question. A weighted logistic regression is the usual first pass:

```{r}
model_data <- hi[essround == 11L & !is.na(poor_health) & !is.na(edu3) &
                   !is.na(income3) & !is.na(age_band)]

fit <- glm(poor_health ~ edu3 + income3 + age_band + factor(gndr) + cntry,
           data = model_data, family = quasibinomial(), weights = anweight)

summary(fit)
```

`quasibinomial()` rather than `binomial()` because the weights are not integer
frequencies, and `binomial()` would emit a warning about non-integer successes
and, worse, report standard errors as though each weighted observation were a
whole one.

Those standard errors are still wrong, though — they assume simple random
sampling, and the ESS is clustered and stratified. Round 11 publishes the design
variables needed to fix that, and `vignette("design-based-estimates")` does the
same model properly with `survey::svyglm()`. Treat the coefficients above as
descriptive and the *p*-values as not yet computed.

## What this vignette worked around

* The module ran in rounds 7 and 11 only. `ess_theme_rounds()` establishes that
  without downloading anything.
* `health` is coded 1 = *Very good* to 5 = *Very bad*. Higher is worse. `hlthhmp`
  codes "No" as 3.
* `eisced` codes 0 and 55 are not flagged missing but are not education levels.
* `hinctnta` is a within-country decile, comparable as a rank and meaningless as
  an amount.
* Five variables use four different missing-code conventions.
* `glm(family = binomial())` with survey weights is the wrong family; the
  standard errors need a design (next vignette).
