---
title: "An index of attitudes to immigration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An index of attitudes to immigration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

> **User story.** As a researcher of public opinion, I need a comparable summary
> measure of attitudes to immigration for the rounds that carried the immigration
> module, so that I can compare countries and compare 2002 with 2014 without the
> comparison being an artefact of how the items were coded.

Catalogue output below is real. Results from survey data are shown as code only.

```{r setup}
library(essurvey2)
library(data.table)
```

## Which rounds

```{r}
ess_theme_rounds("Immigration")[, .(round, description)]
#>    round                                                        description
#>    <int>                                                             <char>
#> 1:     1               ESS round 1 - 2002. Immigration, Citizen involvement
#> 2:     7     ESS round 7 - 2014. Immigration, Social inequalities in health
#> 3:     8 ESS round 8 - 2016. Welfare attitudes, Attitudes to climate change
```

Rounds 1 and 7 ran the module as a headline rotating module — their descriptions
say so. Round 8 is tagged with the theme but its headline modules were welfare and
climate; it carried immigration items inside those. So the module comparison is
rounds 1 and 7, and round 8 can contribute the core items if you want a third
point. The next section shows how to check which, rather than assuming.

## Finding the items, and why the variable group is not enough

Variables are grouped, and the immigration group is the natural place to look:

```{r}
v1 <- ess_variables(1)
v7 <- ess_variables(7)

v1[group_name == "immigration", .N]
#> [1] 58

v7[group_name == "immigration", .N]
#> [1] 35
```

Round 1 documented 58 immigration variables and round 7 documented 35 — the
module was rebuilt, not repeated item for item. Now round 8:

```{r}
v8 <- ess_variables(8)
v8[group_name == "immigration", .(variable_name, variable_label)]
#> Empty data.table (0 rows and 2 cols): variable_name,variable_label
```

Round 8 has **no** `immigration` group at all. Select by group and you would
conclude that round 8 asked nothing about immigration. It did — and so, it turns
out, does round 7 outside its immigration group.

The six items this vignette uses are the long-running core battery: three
admission items and three consequence items. Ask the catalogue where they live in
each round:

```{r}
core <- c("imsmetn", "imdfetn", "impcntr", "imbgeco", "imueclt", "imwbcnt")

rbindlist(lapply(c(1, 7, 8), function(r) {
  v <- ess_variables(r)
  cbind(round = r, v[variable_name %in% core, .(variable_name, group_name)])
}))[, dcast(.SD, variable_name ~ round, value.var = "group_name")]
#>    variable_name           1        7        8
#>           <char>      <char>   <char>   <char>
#> 1:       imbgeco immigration politics politics
#> 2:       imdfetn immigration politics politics
#> 3:       impcntr immigration politics politics
#> 4:       imsmetn immigration politics politics
#> 5:       imueclt immigration politics politics
#> 6:       imwbcnt immigration politics politics
```

All six exist in all three rounds — but they are in the **`immigration` group in
round 1 and the `politics` group in rounds 7 and 8**. The reason is
straightforward once seen: in 2002 these were the rotating module. They worked, so
the ESS promoted them into the core questionnaire. From round 7 the `immigration`
group holds that round's *new* rotating items — the qualification-for-immigration
battery, the aljewlv/almuslv/algyplv items — while the six graduates sit with the
core politics block.

So the group tells you which questionnaire section an item occupied *in that
round*, which is exactly what you want when exploring a module and exactly what
you must not use as a selection rule across rounds. **Groups are for discovery;
names are for selection.** Confirm presence by name:

```{r}
data.table(variable = core,
           r1 = core %in% v1$variable_name,
           r7 = core %in% v7$variable_name,
           r8 = core %in% v8$variable_name)
#>    variable     r1     r7     r8
#>      <char> <lgcl> <lgcl> <lgcl>
#> 1:  imsmetn   TRUE   TRUE   TRUE
#> 2:  imdfetn   TRUE   TRUE   TRUE
#> 3:  impcntr   TRUE   TRUE   TRUE
#> 4:  imbgeco   TRUE   TRUE   TRUE
#> 5:  imueclt   TRUE   TRUE   TRUE
#> 6:  imwbcnt   TRUE   TRUE   TRUE
```

The rotating module's own items are where the group *is* the right tool — for
finding what round 7 asked that round 1 did not:

```{r}
setdiff(v7[group_name == "immigration", variable_name],
        v1[group_name == "immigration", variable_name])
```

## The three ways these items differ from each other

This battery looks uniform and is not. Read all six code lists before combining
anything.

```{r}
ess_variable_info("impcntr", round = 7)$codes
#>     value                            label is_missing
#>    <char>                           <char>     <lgcl>
#> 1:      1 Allow many to come and live here      FALSE
#> 2:      2                       Allow some      FALSE
#> 3:      3                      Allow a few      FALSE
#> 4:      4                       Allow none      FALSE
#> 5:      7                          Refusal       TRUE
#> 6:      8                       Don't know       TRUE
#> 7:      9                        No answer       TRUE
```

```{r}
ess_variable_info("imbgeco", round = 7)$codes
#>      value                label is_missing
#>     <char>               <char>     <lgcl>
#>  1:      0  Bad for the economy      FALSE
#>  2:      1                    1      FALSE
#> ...
#> 11:     10 Good for the economy      FALSE
#> 12:     77              Refusal       TRUE
#> 13:     88           Don't know       TRUE
#> 14:     99            No answer       TRUE
```

Three differences, each of which will silently corrupt an index:

1. **Different ranges.** `imsmetn`, `imdfetn` and `impcntr` run 1–4.
   `imbgeco`, `imueclt` and `imwbcnt` run 0–10. An unweighted mean of the six is
   dominated by the 0–10 items simply because they are wider.

2. **Opposite directions.** On the 1–4 items, 1 is *Allow many* and 4 is *Allow
   none*: **higher means more restrictive**. On the 0–10 items, 0 is *Bad for the
   economy* / *Cultural life undermined* / *Worse place to live* and 10 is the
   positive pole: **higher means more positive**. Average them as they stand and
   the two halves partly cancel out. The resulting index will look reassuringly
   mid-scale and mean nothing.

3. **Different missing codes.** 7/8/9 on the 1–4 items, 77/88/99 on the 0–10
   items:

```{r}
ess_missing_codes(7, variables = core)
#>     variable_name  value      label
#>            <char> <char>     <char>
#>  1:       imsmetn      7    Refusal
#>  2:       imsmetn      8 Don't know
#>  3:       imsmetn      9  No answer
#> ...
#> 10:       imbgeco     77    Refusal
#> 11:       imbgeco     88 Don't know
#> 12:       imbgeco     99  No answer
#> ...
```

That third one is the dangerous one, because a plausible-looking manual recode
gets it wrong in both directions. Treat 7/8/9 as missing on `imbgeco` and you
throw away three legitimate points near the positive end of a 0–10 scale. Treat
77/88/99 as the only missing codes on `impcntr` and every refusal enters your
index as a 7 on a scale whose maximum is 4 — inflating the restrictive end.

Let the default handle it. `recode_missings = TRUE` reads each variable's own
codes from the catalogue:

```{r}
im <- ess_round(c(1, 7), select = c("essround", "cntry", "idno", "anweight", core))
#> ℹ Fetching 2 data files: "ESS1e06_7" and "ESS7e02_3".

im[, lapply(.SD, function(x) c(min = min(x, na.rm = TRUE), max = max(x, na.rm = TRUE))),
   .SDcols = core]
```

Those ranges should be 1–4 for the first three and 0–10 for the rest. Anything
larger means missing codes survived, and the index would be built on them.

### The codes are stable between rounds 1 and 7

Worth confirming, since a rescaling that is right for one round and wrong for the
other is worse than no rescaling:

```{r}
identical(ess_variable_info("impcntr", round = 1)$codes,
          ess_variable_info("impcntr", round = 7)$codes)

identical(ess_variable_info("imbgeco", round = 1)$codes,
          ess_variable_info("imbgeco", round = 7)$codes)
```

## Building the index

Put every item on 0–1, with 1 meaning the most pro-immigration response. Both
transformations then read the same way round.

```{r}
admission <- c("imsmetn", "imdfetn", "impcntr")   # 1..4, higher = more restrictive
consequence <- c("imbgeco", "imueclt", "imwbcnt") # 0..10, higher = more positive

# Reverse and rescale: 1 -> 1.00, 4 -> 0.00
im[, (paste0(admission, "_s")) := lapply(.SD, function(x) (4 - x) / 3),
   .SDcols = admission]

# Rescale only: 0 -> 0.00, 10 -> 1.00
im[, (paste0(consequence, "_s")) := lapply(.SD, function(x) x / 10),
   .SDcols = consequence]

scaled <- paste0(c(admission, consequence), "_s")

im[, lapply(.SD, function(x) round(range(x, na.rm = TRUE), 3)), .SDcols = scaled]
```

Check the reversal did what you meant, on the raw and rescaled columns together —
a sign error here is invisible downstream:

```{r}
im[, .(mean_raw = mean(impcntr, na.rm = TRUE),
       mean_scaled = mean(impcntr_s, na.rm = TRUE))]

im[, .N, keyby = .(impcntr, impcntr_s)]
```

`impcntr == 1` ("Allow many") must map to `impcntr_s == 1`, and `impcntr == 4`
("Allow none") to `0`.

Now the index, with a completeness rule stated rather than implied:

```{r}
im[, n_items := rowSums(!is.na(.SD)), .SDcols = scaled]
im[, pro_immigration := rowMeans(.SD, na.rm = TRUE), .SDcols = scaled]
im[n_items < 4L, pro_immigration := NA_real_]

im[, .N, keyby = n_items]
```

## Does it hold together?

Cronbach's alpha, on complete cases, per round. Six lines of arithmetic and no
extra dependency:

```{r}
cronbach_alpha <- function(dt, cols) {
  m <- as.matrix(dt[, ..cols])
  m <- m[stats::complete.cases(m), , drop = FALSE]
  k <- ncol(m)
  (k / (k - 1)) * (1 - sum(apply(m, 2, stats::var)) / stats::var(rowSums(m)))
}

im[, .(alpha = round(cronbach_alpha(.SD, scaled), 3), n = .N),
   by = essround, .SDcols = scaled]
```

Run it per country as well. An index that is coherent pooled can fall apart in
individual countries, which is a finding about comparability rather than a
nuisance:

```{r}
im[, .(alpha = round(cronbach_alpha(.SD, scaled), 3), n = .N),
   by = .(essround, cntry), .SDcols = scaled][order(alpha)][1:10]
```

Two subscales are often the better answer. Admission and consequence are
conceptually distinct, they were asked with different response formats, and their
correlation is a result worth reporting rather than assuming away:

```{r}
im[, admission_idx := rowMeans(.SD, na.rm = TRUE), .SDcols = paste0(admission, "_s")]
im[, consequence_idx := rowMeans(.SD, na.rm = TRUE), .SDcols = paste0(consequence, "_s")]

im[, .(r = round(cor(admission_idx, consequence_idx, use = "complete.obs"), 3)),
   by = essround]

im[, .(alpha_adm = round(cronbach_alpha(.SD, paste0(admission, "_s")), 3)),
   by = essround, .SDcols = paste0(admission, "_s")]
```

Mixing a 4-point and an 11-point item in one mean gives the wider item more
influence over the total even after rescaling, because it has more variance to
contribute. Rescaling equalises the range, not the variance. If the single index
matters, standardise each item instead — and say which you did.

## Country and round comparison

```{r}
by_country <- im[!is.na(pro_immigration),
  .(n = .N, index = round(weighted.mean(pro_immigration, anweight), 3)),
  by = .(essround, cntry)
]

wide <- dcast(by_country, cntry ~ essround, value.var = "index")
setnames(wide, c("1", "7"), c("r1_2002", "r7_2014"))
wide[!is.na(r1_2002) & !is.na(r7_2014), change := round(r7_2014 - r1_2002, 3)]
wide[order(-change)]
```

Only countries in both rounds can appear in `change`, and there are fewer of them
than either round has:

```{r}
ess_country_rounds("NL")[round %in% c(1, 7), .(round, file_label)]

im[, uniqueN(cntry), by = essround]
wide[!is.na(change), .N]
```

The subscales separately, since they can move in opposite directions — becoming
more positive about immigration's effects while wanting less of it is a coherent
position and a real pattern:

```{r}
im[!is.na(admission_idx) & !is.na(consequence_idx),
  .(admission = round(weighted.mean(admission_idx, anweight), 3),
    consequence = round(weighted.mean(consequence_idx, anweight), 3)),
  by = .(essround)
]
```

As in the other vignettes, these are point estimates without design-based
standard errors. Rounds 1 and 7 are also the awkward case for that: round 7's
design variables need the SDDF file and round 1 has none published through this
API. See `vignette("design-based-estimates")`.

## Adding round 8

The core items are in round 8 even though the group is not, so a third point is
one argument away:

```{r}
im3 <- ess_round(c(1, 7, 8), select = c("essround", "cntry", "idno", "anweight", core))
```

Before pooling, verify the codes did not shift:

```{r}
lapply(core, function(v) {
  c8 <- ess_variable_info(v, round = 8)$codes
  c7 <- ess_variable_info(v, round = 7)$codes
  data.table(variable = v, identical = identical(c7, c8))
}) |> rbindlist()
```

Report round 8 as the core items only, not as the module. It is not the same
instrument, and the surrounding questions differ — which affects answers.

## What this vignette worked around

* The six core items are in the `immigration` group in round 1 and the `politics`
  group in rounds 7 and 8, where the `immigration` group holds that round's new
  rotating items instead. Round 8 has no `immigration` group at all. Groups are
  for discovery; names are for selection.
* The battery mixes 1–4 and 0–10 response formats.
* The two halves run in opposite directions: higher is *more restrictive* on the
  admission items and *more positive* on the consequence items.
* Missing codes differ within the battery — 7/8/9 against 77/88/99 — so a
  hand-written recode gets one half wrong whichever convention it picks. The
  default `recode_missings = TRUE` reads them per variable.
* Rescaling equalises range, not variance. A 4-point and an 11-point item do not
  contribute equally to a mean of the two.
* Only countries present in both rounds can enter a change score.
