---
title: "Recording exactly which ESS data you analysed"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recording exactly which ESS data you analysed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

> **User story.** As the corresponding author of a paper using ESS data, I need a
> record of exactly which data files, editions and DOIs produced my results, so
> that I can cite them as the ESS asks and so that a reader can tell whether a
> failure to reproduce my numbers is their mistake, my mistake, or a new edition.

Catalogue output below is real, from a session run in July 2026. Your own run will
differ, and the point of this vignette is that the difference is knowable.

```{r setup}
library(essurvey2)
library(data.table)
```

## The problem with resolving DOIs at run time

`essurvey2` hard-codes nothing. `ess_round(11)` asks the catalogue which data
file is round 11's integrated file, and downloads whatever it is told. That is
what makes a newly published round or a corrected edition appear without a package
update — and it is also what makes the script's output a moving target.

Editions are revised regularly. Every round in the series is past its first:

```{r}
ess_rounds()[, .(round, file_label, edition)]
#>     round file_label edition
#>     <int>     <char>  <char>
#>  1:     1  ESS1e06_7     6.7
#>  2:     2  ESS2e03_6     3.6
#>  3:     3  ESS3e03_7     3.7
#>  4:     4  ESS4e04_6     4.6
#>  5:     5  ESS5e03_6     3.6
#>  6:     6  ESS6e02_7     2.7
#>  7:     7  ESS7e02_3     2.3
#>  8:     8  ESS8e02_3     2.3
#>  9:     9  ESS9e03_3     3.3
#> 10:    10 ESS10e03_3     3.3
#> 11:    11 ESS11e04_2     4.2
```

Round 1's data has been through six editions and seven sub-releases. Each edition
has its own DOI, and a new one may add a country, correct a weight or fix a
recode. So the same script, run a year apart, can legitimately produce different
numbers. Without a record you cannot tell that from a bug.

The fix is not to hard-code DOIs — that puts you on a stale edition indefinitely,
and a DOI in a script gives no signal when a better one exists. The fix is to
resolve the current edition, use it, and **write down what you got**.

## A provenance record

Everything needed is in the catalogue, and none of it requires downloading data.
`ess_data_file_info()` gives the per-file detail:

```{r}
f <- ess_data_files(11, kind = "integrated")
info <- ess_data_file_info(f$datafile_id, f$datafile_version)

str(info, max.level = 1)
#> Classes 'data.table' and 'data.frame':	1 obs. of  11 variables:
#>  $ file_label      : chr "ESS11e04_2"
#>  $ description     : chr "ESS11 - integrated file, edition 4.2"
#>  $ edition         : chr "4.2"
#>  $ doi             : chr "10.21338/ess11e04_2"
#>  $ doi_url         : chr "https://doi.org/10.21338/ess11e04_2"
#>  $ published       : chr "2026-07-02"
#>  $ n_variables     : int 709
#>  $ n_countries     : int 30
#>  $ countries       :List of 1
#>  $ datafile_id     : chr "242aaa39-3bbb-40f5-98bf-bfb1ce53d8ef"
#>  $ datafile_version: int 179
```

`published` is the date that edition was released, which is the field that tells a
future reader whether a newer one exists. `n_countries` and `countries` record the
coverage you actually had:

```{r}
info$n_countries
#> [1] 30

info$countries[[1]]
#> [1] "AT" "BE" "BG" "HR" ...
```

A helper worth keeping in a project's `R/` directory. It takes the rounds an
analysis used and returns one row per file:

```{r}
ess_provenance <- function(rounds, kind = "integrated") {
  files <- ess_data_files(rounds = rounds, kind = kind)

  rbindlist(lapply(seq_len(nrow(files)), function(i) {
    info <- ess_data_file_info(files$datafile_id[[i]], files$datafile_version[[i]])
    data.table(
      round = files$round[[i]],
      file_label = info$file_label,
      edition = info$edition,
      doi = info$doi,
      published = info$published,
      n_variables = info$n_variables,
      n_countries = info$n_countries,
      countries = paste(sort(info$countries[[1]]), collapse = " ")
    )
  }))
}

prov <- ess_provenance(c(7, 11))
prov[, .(round, file_label, edition, doi, published, n_countries)]
```

Write it out next to the results, and record the environment as well. The data is
half of what makes a number reproducible; the code that read it is the other half.

Everything below writes into `out_dir`. This vignette points it at the session's
temporary directory so that running the code leaves nothing behind; in your own
project, point it at the output folder you archive with the paper:

```{r}
out_dir <- tempdir()          # your project:  out_dir <- "output"
prov_csv <- file.path(out_dir, "ess-provenance.csv")

fwrite(prov, prov_csv)

writeLines(
  c(
    paste("essurvey2:", as.character(packageVersion("essurvey2"))),
    paste("arrow:    ", as.character(packageVersion("arrow"))),
    paste("R:        ", R.version.string),
    paste("run:      ", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"))
  ),
  file.path(out_dir, "ess-session.txt")
)
```

## One rough edge to know about

The `doi` column of `ess_data_files()` is passed through from the catalogue
without normalisation, and the catalogue is not entirely consistent. As of
July 2026 one file of the 64 — round 7's alcohol module — carries its DOI as a
URL, with a leading space:

```{r}
all_files <- ess_data_files()
odd <- all_files[!grepl("^10\\.", doi)]
odd[, .(round, file_label, doi)]
#>    round file_label                                  doi
#>    <int>     <char>                               <char>
#> 1:     7 ESS7ALCe02  https://doi.org/10.21338/ess7alce02

nchar(odd$doi)
#> [1] 36
```

Downloading is unaffected — `ess_data()` normalises the DOI it is given, stripping
a `https://doi.org/` prefix and surrounding whitespace, so the value works as it
stands:

```{r}
ess_parse_doi(" https://doi.org/10.21338/ess7alce02")
#> $prefix
#> [1] "10.21338"
#>
#> $suffix
#> [1] "ess7alce02"
```

What it breaks is anything that treats `doi` as a key or prints it as a citation.
An exact-match filter silently returns nothing:

```{r}
all_files[doi == "10.21338/ess7alce02"]                  # zero rows
all_files[trimws(doi) %like% "ess7alce02$"]              # one row
```

So normalise on the way into a provenance table, and the citation is clean and the
joins work:

```{r}
canonical_doi <- function(x) {
  x <- trimws(x)
  sub("^https?://(dx\\.)?doi\\.org/", "", x, ignore.case = TRUE)
}

prov[, doi := canonical_doi(doi)]
prov[, doi_url := paste0("https://doi.org/", doi)]
```

The same caution applies to the round 8 duplicate that `ess_data_files()` warns
about — two files sharing one DOI, which no amount of local tidying can resolve:

```{r}
ess_data_files(8)
#> Warning: The ESS catalogue gives 2 data files the same DOI "10.21338/ess8timee01".
#> ✖ "ESS8TIMEe01" and "ESS8CFe03"
#> ℹ Downloading it returns one file, which may not be the one you asked for.
#> ℹ This is an error in the catalogue, not in essurvey2; report it to 'essdatasupport@sikt.no'.
```

If your analysis touches either file, record the warning alongside the provenance
table and check the `name` column of what actually arrived. The package reports the
ambiguity rather than picking a side, because picking wrong would hand you the
wrong data without saying so.

## Detecting an edition change

Having written a provenance record, use it. Re-resolve the catalogue and compare
against the file on disk:

```{r}
check_provenance <- function(path) {
  old <- fread(path, colClasses = "character")
  new <- ess_provenance(as.integer(old$round))

  cmp <- merge(
    old[, .(round = as.integer(round), old_edition = edition, old_doi = doi)],
    new[, .(round, new_edition = edition, new_doi = doi)],
    by = "round"
  )

  cmp[, changed := old_edition != new_edition]
  cmp
}

check_provenance(prov_csv)
```

`changed == TRUE` means the ESS has published a new edition since your run. That
is not an error — it is the signal to re-run, compare, and decide whether the
results moved enough to matter.

Catalogue answers are memoised for the session, so a long-running session will not
notice a mid-session publication. `ess_catalogue_refresh()` forgets them:

```{r}
ess_catalogue_refresh()
ess_rounds()[, .(round, file_label, edition)]
```

## The cache is not a provenance record

Downloads are cached on disk, keyed by DOI, format and missing-value treatment:

```{r}
ess_cache_list()
ess_cache_dir()
```

The cache is why the second run of a script is fast, and it does keep the file that
produced your results. But it is not a substitute for the provenance table, for
two reasons.

First, the cache key is the DOI, and a new edition has a new DOI. So a new edition
does not overwrite the old file — it is downloaded alongside it. Your cache can
hold both `10.21338/ess11e04_2` and its successor, and the filename is the only
thing distinguishing them:

```{r}
ess_cache_list()[, .(file, size)]
#>                                    file    size
#> 1: 10.21338_ess11e04_2__recoded.parquet 12.4 MB
```

Second, `ess_cache_clear()` exists and gets used, and a cache directory is not
something you would archive with a paper. Keep the CSV.

If you want the exact file kept with the project rather than in a cache under the
user's home directory, download it explicitly. Again `out_dir` stands in for the
folder you archive — swap in `"data-raw"` or wherever the project keeps its inputs:

```{r}
parquet <- file.path(out_dir, "ess11e04_2.parquet")
ess_download_file("10.21338/ess11e04_2", path = parquet)

# Re-read it later without touching the network.
dt <- ess_read_file(parquet, select = c("essround", "cntry", "anweight"))
```

Or move the whole cache somewhere the project controls:

```{r}
options(essurvey2.cache_dir = file.path(out_dir, "ess-cache"))
```

## Citing the data

The ESS asks that the data files an analysis uses be cited. The provenance table
holds the fields a citation needs — file label, edition, DOI and publication date:

```{r}
prov[, sprintf("European Social Survey (%s). %s, ed. %s. Sikt. https://doi.org/%s",
               substr(published, 1, 4), file_label, edition, doi)]
```

Check the ESS's current citation requirements at <https://ess.sikt.no> before
submitting; the fields above are what this package can tell you, not a claim about
the required format, which the ESS sets and revises.

Cite the *data files*, not the study. The API only serves data file DOIs, and
study-level DOIs are rejected — which is a useful reminder of which kind you are
holding:

```{r}
tryCatch(
  ess_data("10.21338/NSD-ESS10-2020"),
  essurvey2_error_doi = function(e) conditionMessage(e)
)
```

## What this vignette worked around

* Nothing is hard-coded, so the edition a script gets depends on when it runs.
  Record what you got; do not pin a DOI.
* `published` in `ess_data_file_info()` is what dates your extract.
* One of the 64 data file DOIs is served as a URL with a leading space. Downloads
  cope; exact-match filters and citations do not. Normalise it.
* Round 8 has two data files sharing one DOI. The package warns; the ambiguity is
  upstream.
* The cache keeps files but is not an archive — a new edition lands beside the old
  one rather than replacing it, and `ess_cache_clear()` exists.
* Catalogue answers are memoised per session; `ess_catalogue_refresh()` forgets
  them.
