---
title: "Trends in institutional trust across countries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trends in institutional trust across countries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

> **User story.** As a comparative political sociologist, I need a weighted
> country-by-round series of trust in national institutions covering the whole
> ESS, so that I can describe trends that are comparable across countries and
> defensible under the ESS's own weighting guidance.

Nothing in this vignette is evaluated when the package is built, because every
download needs a user ID. The catalogue output shown below is real, copied from a
session run against the live catalogue. Results computed from survey data are
deliberately *not* pasted in: ESS editions are revised, and a number printed here
would quietly stop matching what you get. Run the code and read your own numbers.

```{r setup}
library(essurvey2)
library(data.table)
```

## The question

Has trust in national political institutions fallen across Europe since 2002,
and does the answer depend on which countries you count?

That second clause is the whole difficulty. The ESS is not a panel of countries:
39 have taken part at least once, and only eleven have been in every round. A
raw European average over rounds therefore mixes a change in attitudes with a
change in who was asked.

## Finding the items

Trust is part of the *Politics* core module, so it is present in every round.
`ess_theme_rounds()` confirms that before you download anything:

```{r}
ess_theme_rounds("Politics")[, .(round)]
#>     round
#>  1:     1
#>  2:     2
#> ...
#> 11:    11
```

Variables in an integrated file are organised into groups, which is usually a
quicker way in than searching labels:

```{r}
vars <- ess_variables(11)
unique(vars[, .(group_name, group_label)])
#>      group_name                                                      group_label
#>          <char>                                                           <char>
#>  1:   mandatory                                  Identifier and weight variables
#>  2:       media                                           Media and social trust
#>  3:    politics                                                         Politics
#> ...
```

The trust battery itself:

```{r}
vars[grepl("^trst", variable_name), .(variable_name, variable_label)]
#>    variable_name                   variable_label
#>           <char>                           <char>
#> 1:       trstprl    Trust in country's parliament
#> 2:       trstlgl        Trust in the legal system
#> 3:       trstplc              Trust in the police
#> 4:       trstplt             Trust in politicians
#> 5:       trstprt       Trust in political parties
#> 6:        trstep Trust in the European Parliament
#> 7:        trstun      Trust in the United Nations
```

All seven are 0–10 scales with the same missing codes, which is worth checking
rather than assuming — batteries that look uniform often are not (see
`vignette("immigration-attitudes")` for one that is not):

```{r}
ess_variable_info("trstprl", round = 11)$codes
#>      value           label is_missing
#>     <char>          <char>     <lgcl>
#>  1:      0 No trust at all      FALSE
#>  2:      1               1      FALSE
#> ...
#> 11:     10  Complete trust      FALSE
#> 12:     77         Refusal       TRUE
#> 13:     88      Don't know       TRUE
#> 14:     99       No answer       TRUE
```

### One item is not in round 1

`trstprt` — trust in political parties — was added in round 2. Checking the
whole series before building an index costs one catalogue call per round and
saves a misleading trend:

```{r}
items <- c("trstprl", "trstlgl", "trstplc", "trstplt", "trstprt")

availability <- rbindlist(lapply(1:11, function(r) {
  data.table(round = r, variable_name = items,
             present = items %in% ess_variables(r)$variable_name)
}))

dcast(availability, variable_name ~ round, value.var = "present")
#> Key: <variable_name>
#>    variable_name     1    2    3    4    5    6    7    8    9   10   11
#> 1:       trstlgl  TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE
#> 2:       trstplc  TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE
#> 3:       trstplt  TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE
#> 4:       trstprl  TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE
#> 5:       trstprt FALSE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE
```

`ess_round()` stacks rounds with `fill = TRUE`, so round 1 would simply carry
`NA` for `trstprt`. A row mean over the five items would then rest on four items
in round 1 and five thereafter — a discontinuity manufactured by the analyst,
not by respondents. Two honest options: drop `trstprt`, or start at round 2.
This vignette drops it and keeps all eleven rounds.

## Which weight

Round 11's mandatory group carries four weights, and the choice between them
matters more than most of the modelling that follows:

```{r}
vars[grepl("wght|weight", variable_name), .(variable_name, variable_label)]
#>    variable_name                                                    variable_label
#>           <char>                                                            <char>
#> 1:       dweight                                                     Design weight
#> 2:       pspwght                Post-stratification weight including design weight
#> 3:       pweight Population size weight (must be combined with dweight or pspwght)
#> 4:      anweight                                                   Analysis weight
#> 5:       weighta                                         Weight of respondent (kg)
```

Five hits, four weights. `weighta` is the respondent's body weight in kilograms,
from the health module — a reminder that a name pattern is a search, not a
selection, and that reading the labels is not optional. The variable group is the
safer filter here: the four survey weights are all in `mandatory`, and `weighta`
is not.

The catalogue documents `anweight` for itself, and the wording settles the
question:

```{r}
ess_variable_info("anweight", round = 11)$description
#> [1] "This weight is suitable for all types of analysis, including when you are
#>      studying just one country, when you compare across countries, or when you
#>      are studying groups of countries. For more information, please see the
#>      document \"Weighting ESS data\" (User guide).
#>      Ukraine: Post-stratification adjustment has been set to 1.00 for all
#>      cases. For more information see section on Country documentation."
```

So: use `anweight`, and it is available in all eleven rounds. Two footnotes.

First, the catalogue's `role` field still marks `pspwght` as the default weight,
which predates `anweight`:

```{r}
vars[role == "DefaultWeight", .(variable_name, variable_label)]
#>    variable_name                                     variable_label
#>           <char>                                             <char>
#> 1:       pspwght Post-stratification weight including design weight
```

Take `anweight`'s own description over that tag. `pspwght` alone weights every
country to the same total, which is what you want *within* a country and wrong
the moment you pool.

Second, `anweight` is the post-stratification weight scaled by the population
size weight. You do not have to take that on trust — check it on the data you
downloaded:

```{r}
dt <- ess_round(11, select = c("anweight", "pspwght", "pweight"))
dt[, summary(anweight - pspwght * pweight)]
```

## Downloading

Five items, a country code, a round number and a weight: nine columns out of the
600–700 each round holds. Because parquet is columnar, `select` is the difference
between a few megabytes and a few hundred.

```{r}
items <- c("trstprl", "trstlgl", "trstplc", "trstplt")

trust <- ess_round(
  1:11,
  select = c("essround", "cntry", "idno", "anweight", items)
)
#> ℹ Fetching 11 data files: "ESS1e06_7", "ESS2e03_6", … and "ESS11e04_2".

dim(trust)
```

The first call downloads; every call after that reads from the cache, in this
session or in six months' time:

```{r}
ess_cache_list()[, .(file, size)]
```

## An index, and what to do about partial answers

```{r}
trust[, n_answered := rowSums(!is.na(.SD)), .SDcols = items]
trust[, trust_index := rowMeans(.SD, na.rm = TRUE), .SDcols = items]

trust[, .N, by = n_answered][order(n_answered)]
```

A row mean with `na.rm = TRUE` averages whatever is there, so a respondent who
answered one item of four gets an index built from that one item. Require a
minimum instead, and say what you required:

```{r}
trust[n_answered < 3L, trust_index := NA_real_]
```

## The country composition problem

`ess_countries()` reports how many rounds each country appears in:

```{r}
countries <- ess_countries()
countries[, .N, by = n_rounds][order(-n_rounds)]

balanced <- countries[n_rounds == 11L, country_code]
balanced
#>  [1] "BE" "FI" "FR" "HU" "IE" "NL" "NO" "PT" "SI" "CH" "GB"
```

Eleven countries of 39 have been in every round. Now the two series side by
side — everyone who happened to be in each round, against the fixed eleven:

```{r}
setkey(trust, essround)

series <- trust[
  !is.na(trust_index),
  .(all_participants = weighted.mean(trust_index, anweight)),
  by = essround
]

series[
  trust[cntry %in% balanced & !is.na(trust_index),
        .(balanced_panel = weighted.mean(trust_index, anweight)),
        by = essround],
  on = "essround"
][order(essround)]
```

Where the two columns diverge, the divergence is composition, not opinion. Report
the balanced panel, or report both.

## Country trends

Per country and round, with the number of respondents behind each figure:

```{r}
by_country <- trust[
  !is.na(trust_index),
  .(n = .N, trust = round(weighted.mean(trust_index, anweight), 2)),
  by = .(cntry, essround)
]

dcast(by_country[cntry %in% balanced], cntry ~ essround, value.var = "trust")
```

First round to last, for countries observed at both ends:

```{r}
change <- by_country[
  , .(first = essround[which.min(essround)], last = essround[which.max(essround)],
      from = trust[which.min(essround)], to = trust[which.max(essround)],
      n_rounds = .N),
  by = cntry
][n_rounds >= 8L]

change[, delta := round(to - from, 2)]
change[order(delta), .(cntry, first, last, from, to, delta)]
```

Two cautions on that table. The endpoints are different calendar years for
different countries, since not every country is in every round. And a difference
of means says nothing about whether it is distinguishable from noise — for that
you need standard errors that respect the ESS's clustered sample, which is
`vignette("design-based-estimates")`.

## Recording what you used

Editions are revised. The trend you just computed belongs to a specific set of
files, and `ess_rounds()` names them:

```{r}
ess_rounds()[, .(round, file_label, edition)]
#>     round file_label edition
#>     <int>     <char>  <char>
#>  1:     1  ESS1e06_7     6.7
#>  2:     2  ESS2e03_6     3.6
#> ...
#> 11:    11 ESS11e04_2     4.2
```

`vignette("reproducible-provenance")` turns that into a citation table to keep
beside the results.

## What this vignette worked around

* `trstprt` is absent from round 1, and stacking would have hidden that behind
  `NA`. Checking availability across rounds is cheap; discovering it in a
  referee report is not.
* `anweight` is the weight for this job, in every round. `pspwght` is still
  tagged `DefaultWeight` in the catalogue.
* A European mean over rounds confounds attitude change with country
  composition. The balanced panel of eleven is one defensible answer.
* `rowMeans(na.rm = TRUE)` silently accepts one item out of four. Set a
  threshold.
